### 2024-09-05 parts/intro.tex
putting things in appendix

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\NPoly$, for \kl{polyregular functions}
with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$ for
\kl{star-free polyregular functions} with unary output) \cite{DOUE21,DOUE22}.
From this point of view, $\NPoly$ is a subclass of a larger and well-studied
class of functions called (non-commutative) \kl{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural (strict) subclass of
\kl{$\Nat$-rational series} ($\NRat$), another model of computation that aims
at generalizing the notion of regular language to functions with outputs in a
\emph{commutative} semiring \cite{BERE88,BERE10}.\footnote{In particular, the
    notion of \kl{rational series} with output in $\Gamma^*$ is not well-defined.} It was frequently rediscovered in the literature that
\kl{$\Nat$-polyregular functions} (although not often called this way) are a
subclass of \kl{rational series} \cite{SCHU62,KRRC13,CDTL23}.

@@

\AP This approach through rational series has several limitations. First,
$\ZPoly$ does not correspond to a restriction of \kl{polyregular functions},
because of the extra expressive power introduced by the negative outputs
\cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
is also an open problem (see e.g. \cite{KARH77}), that remains open even in the
restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It
was conjectured that, being ``computable using coefficients in $\Nat$'' and being
``star-free'' were two orthogonal properties, and therefore that the following
equation would hold: $\NSF = \NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}.
This would imply the decidability of aperiodicity for single letter output
polyregular functions since aperiodicity is decidable for $\ZSF$. We refer the
reader to \cref{previously-known-inclusions:fig} to get an overview of the
different classes of functions appearing in the paper.

@@

\subparagraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}, that is where the output of
the functions remains unchanged when the letters of the input word are
permuted. These functions from $\Sigma^*$ to $\Rel$ can be understood as
functions from $\Nat^{\card{\Sigma}}$ to $\Rel$, and are therefore easier to
study using classical mathematical tools. The simplest kind of functions in
this class are actual polynomial functions, which were studied by Karhumäki in
\cite{KARH77}. Our first contribution is to provide a counter-example to the
main theorem of \cite{KARH77} that claimed to characterize the (multivariate)
polynomials in $\Rat[\vec{X}]$ that can be computed by some \kl{$\Nat$-rational
series} (\cref{thm:counter-example}). Our second contribution is to correct
this error, and provide (effective) characterizations of polynomials in
$\Rat[\vec{X}]$ that can be computed by \kl{$\Rel$-polyregular functions}
(\cref{integer-binomial-polynomials:cor}), and by \kl{$\Nat$-polyregular
functions} (\cref{decide-rat-poly-npoly:cor}). Our third contribution
of this paper is to answer \cite[Conjecture 7.61]{DOUE23} (stating that $\NSF =
\NPoly \cap \ZSF$) positively in the \kl{commutative} case, and prove that the
membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
decidable, with effective conversion algorithms. This effectively shows that
all arrows in the upper left square of \cref{previously-known-inclusions:fig}
are decidable in the \kl{commutative} setting.

>>>>

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\NPoly$, for \kl{polyregular functions}
with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$ for
\kl{star-free polyregular functions} with unary output) \cite{DOUE21,DOUE22}.
From this point of view, $\NPoly$ is a subclass of a larger and well-studied
class of functions called (non-commutative) \kl{rational series}
\cite{REUT80,BERE88,BERE10}. This other model of computation generalizes the
notion of regular language to functions with outputs in a \emph{commutative}
semiring, and it was frequently rediscovered that \kl{$\Nat$-polyregular
functions} form a (strict) subclass of \kl{rational series} with outputs in
$\Nat$, called \kl{$\Nat$-rational series} ($\NRat$)
\cite{SCHU62,KRRC13,CDTL23}.

@@

\AP This approach through rational series bears numerous limitations. First,
$\ZPoly$ does not correspond to a restriction of \kl{polyregular functions},
because of the extra expressive power introduced by the negative outputs
\cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
is an open problem (see e.g. \cite{KARH77}), that remains open even in the
simple case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It was
conjectured that, being ``computable using coefficients in $\Nat$'' and being
``star-free'' were two orthogonal properties, and therefore that the following
equation would hold: $\NSF = \NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}.
This would imply the decidability of aperiodicity for single letter output
polyregular functions since aperiodicity is decidable for $\ZSF$. We refer the
reader to \cref{previously-known-inclusions:fig} to get an overview of the
different classes of functions appearing in the paper.

@@

\subparagraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}, that is where the output of
the functions remains unchanged when the letters of the input word are
permuted. These functions from $\Sigma^*$ to $\Rel$ can be understood as
functions from $\Nat^{\card{\Sigma}}$ to $\Rel$, and are therefore easier to
study using classical mathematical tools. The simplest kind of functions in
this class are actual (multivariate) polynomial functions, which were studied
by Karhumäki in \cite{KARH77}. Our analysis of \kl{commutative}
\kl{$\Nat$-polyregular functions} is divided in two main parts: the first one
completely revisits the results of \cite{KARH77} on polynomials, and the second
one extends the former decidability results to arbitrary \kl{commutative}
\kl{$\Nat$-polyregular functions}.

Our first contribution is to provide a counter-example to the main theorem of
\cite{KARH77} that claimed to characterize the (multivariate) polynomials in
$\Rat[\vec{X}]$ that can be computed by some \kl{$\Nat$-rational series}
(\cref{thm:counter-example}). Our second contribution is to correct this error,
and provide (effective) characterizations of polynomials in $\Rat[\vec{X}]$
that can be computed by \kl{$\Rel$-polyregular functions}
(\cref{integer-binomial-polynomials:cor}), and by \kl{$\Nat$-polyregular
functions} (\cref{decide-rat-poly-npoly:cor}). These characterizations also
show that polynomials computable using $\NRat$ (resp. $\ZRat$) are also
computable using $\NSF$ (resp. $\ZSF$), that is: polynomials are inherently
\emph{star-free}. Furthermore, these characterizations provide effective
descriptions of polynomials that can be expressed in $\ZPoly$ as those obtained
using integer combinations of products of \emph{binomial coefficients} (called
\kl{integer binomial polynomials}, defined page \kpageref{integer binomial
polynomial}) and similarly for $\NPoly$ using the notion of \kl{strongly
natural binomial polynomials} (defined page \kpageref{strongly natural binomial
polynomial}).

Our third and last contribution
of this paper is to answer \cite[Conjecture 7.61]{DOUE23} (stating that $\NSF =
\NPoly \cap \ZSF$) positively in the \kl{commutative} case, and prove that the
membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
decidable, with effective conversion algorithms. This effectively shows that
all arrows in the upper left square of \cref{previously-known-inclusions:fig}
are decidable in the \kl{commutative} setting.
